Generating Interactive WebSocket Applications in TypeScript

### 2020-03-09 PLACES_2020/encoding.tex
Save some space

We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
this section.
For server-side targets, we encode EFSM states into TypeScript types and
consider branching (receiving) and selection (sending) states separately.
We assign TypeScript encodings of states to their state identifiers in the
EFSM, providing syntactic sugar when referring to the successor state when
encoding the current state.
For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
encoding as $\llbracket S \rrbracket$.
We outline the encoding below using examples from the
\textit{Noughts and Crosses} game (\cref{lst:svr}).

@@

Components can also render other components, which gives rise
to a parent/child relationship between components.
Parents can pass data to children as \textit{props} (short for properties).
Going back to the aforementioned example, the counter component could
render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
its \texttt{render()} function, propagating the \texttt{count} from its state
to the child.
This enables reusability, and for our use case, gives control to the parent
on what data to pass to its children (e.g. pass the payload of a received
message to a child to render).

@@

  \captionof{figure}{EFSM for \texttt{P1}.}
  \label{fig:efsmp1}
\end{wrapfigure}

We refer to the \texttt{P1} EFSM (\cref{fig:efsmp1}) as a running example in
this section.
Preserving behavioural typing and channel linearity is challenging
for browser-side applications due to EFSM transitions being triggered by user
events:
in the case of \textit{Noughts and Crosses}, once the user makes a move by
clicking on a cell on the game board, this click event must be deactivated
until the user's next turn, otherwise the user can click again and violate
channel linearity.
Our design goal is to enforce this statically through the generated APIs.

For browser-side targets, we extend the approach presented in \cite{MVU2019} on
\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
architecture.
% FEEDBACK: introduce MVU and model type
An MVU application features a \textit{model} encapsulating application
state, a \textit{view function} rendering the state on the Document Object Model (DOM), and an
\textit{update function} handling \textit{messages} produced by the
rendered model to produce a new model.
The concept of model types express type dependencies between these
components: a \emph{model type} uniquely defines a \textit{view function},
set of \textit{messages} and \textit{update function} -- rather than
producing a new model, the update function defines valid transitions to
other model types.
We leverage the correspondence between model types and states in the EFSM:
each state in the EFSM is a model type, the set of messages represent
the possible (IO) actions available at that state,
and the update function defines which successor state to transition to,
given the supported IO actions at this state.

@@

\paragraph{Model transitions}
Transitions are encoded as React component props onto the encoded states by the
session runtime (\cref{section:clientruntime}).
We motivate the design choice of not exposing channel resources to provide
guarantees on channel usage.
React components in TypeScript are
\textit{generic} \cite{TypeScriptSpec}, parameterised by the permitted
types of prop and state.
The parameters allow us to leverage the TypeScript compiler to
verify that the props for model transitions stay local to the state they are
defined for.
The model transitions for EFSMs are message send and receive.

\subparagraph{Sending}
We make the assumption that message sending is triggered by
some user-driven UI event (e.g. clicking a button, pressing a key on the
keyboard) which interacts with some DOM element.
We could pass a
\texttt{send()} function as a prop to the sending state, but the developer
would be free to call the function multiple times which makes channel reuse
possible.
Instead, we pass a \textit{factory function} as a prop, which will,
given an HTML event and an event handler function, return a fresh React
component that binds the sending action on construction.
So once the bound event is triggered, our session runtime executes the event
handler function to obtain the payload to send, perform the send
\textit{exactly once} and transition to (which, in practice, means render) the
successor state.

@@

The session runtime can be interpreted as an abstraction on top of the React
VDOM that implements the EFSM by construction.
The session runtime itself is a React component too, named after the endpoint
role identifier:
it opens the WebSocket connection to the server, keeps track of the current
EFSM state as part of its React component state, and most importantly, renders
the React component encoding of the active EFSM state.
Channel communications are managed by the runtime, which allows it to render
the successor of a receive state upon receiving a message from the channel.
Similarly, the session runtime is responsible for passing the required props
for model transitions to EFSM state React components.
The session runtime component is rendered by the developer and requires, as
props, the \textit{endpoint URL} (so it can open the connection) and a list of
\textit{concrete state components}.

The developer writes their own implementation of each state (mainly to
customise how the state is rendered and inject business logic into state
transitions) by extending the abstract React class components.
The session runtime requires references to these concrete components in order to
render the user implementation accordingly.

\subsubsection{Affine Channel Usage}
A limitation of our browser-side session type encoding is only being able to
guarantee that channel resources are used \textit{at most once} as opposed to
\textit{exactly once}.

>>>>

We refer to the \texttt{Svr} EFSM (\cref{fig:efsmsvr}) as a running example in
this subsection.
For server-side targets, we encode EFSM states into TypeScript types and
consider branching (receiving) and selection (sending) states separately.
We assign TypeScript encodings of states to their state identifiers in the
EFSM, providing syntactic sugar when referring to the successor state when
encoding the current state.
For any state $S$ in the EFSM, we refer to the TypeScript type alias of its
encoding as $\llbracket S \rrbracket$.
We outline the encoding below using examples from the
\textit{Noughts and Crosses} game (\cref{lst:svr}).

@@

Components can also render other components, which gives rise
to a parent/child relationship between components.
Parents can pass data to children as \textit{props} (short for properties).
Going back to the aforementioned example, the counter component could
render a child component \texttt{<StyledDiv count=\{this.state.count\} />} in
its \texttt{render()} function, propagating the \texttt{count} from its state
to the child.
This enables reusability, and for our use case, gives control to the parent
on what data to pass to its children (e.g.\ pass the payload of a received
message to a child to render).

@@

  \captionof{figure}{EFSM for \texttt{P1}.}
  \label{fig:efsmp1}
  \vspace{-5mm}
\end{wrapfigure}

We refer to the \texttt{P1} EFSM (\cref{fig:efsmp1}) as a running example in
this subsection.
Preserving behavioural typing and channel linearity is challenging
for browser-side applications due to EFSM transitions being triggered by user
events:
in the case of \textit{Noughts and Crosses}, once the user makes a move by
clicking on a cell on the game board, this click event must be deactivated
until the user's next turn, otherwise the user can click again and violate
channel linearity.
Our design goal is to enforce linearity statically through the generated APIs.

For browser-side targets, we extend the approach presented in \cite{MVU2019} on
\textit{multiple model types} motivated by the \textit{Model-View-Update} (MVU)
architecture.
% FEEDBACK: introduce MVU and model type
An MVU application features a \textit{model} encapsulating application
state, a \textit{view function} rendering the state on the Document Object Model (DOM), and an
\textit{update function} handling \textit{messages} produced by the
rendered model to produce a new model.
The concept of model types express type dependencies between these
components: a \emph{model type} uniquely defines a \textit{view function},
set of \textit{messages} and \textit{update function} --- rather than
producing a new model, the update function defines valid transitions to
other model types.
We leverage the correspondence between model types and states in the EFSM:
each state in the EFSM is a model type, the set of messages represent
the possible (IO) actions available at that state,
and the update function defines which successor state to transition to,
given the supported IO actions at this state.

@@

\paragraph{Model Transitions}
Transitions are encoded as React component props onto the encoded states by the
session runtime (\cref{section:clientruntime}).
We motivate the design choice of not exposing channel resources to provide
guarantees on channel usage.
React components in TypeScript are
\textit{generic} \cite{TypeScriptSpec}, parameterised by the permitted
types of prop and state.
The parameters allow us to leverage the TypeScript compiler to
verify that the props for model transitions stay local to the state they are
defined for.
The model transitions for EFSMs are message send and receive.

\subparagraph{Sending}
We make the assumption that message sending is triggered by
some user-driven UI event (e.g.\ clicking a button, pressing a key on the
keyboard) which interacts with some DOM elements.
We could pass a
\texttt{send()} function as a prop to the sending state, but the developer
would be free to call the function multiple times which makes channel reuse
possible.
Instead, we pass a \textit{factory function} as a prop, which will,
given an HTML event and an event handler function, return a fresh React
component that binds the sending action on construction.
So once the bound event is triggered, our session runtime executes the event
handler function to obtain the payload to send, perform the send
\textit{exactly once} and transition to (which, in practice, means render) the
successor state.

@@

The session runtime can be interpreted as an abstraction on top of the React
VDOM that implements the EFSM by construction.
The session runtime itself is a React component too, named after the endpoint
role identifier:
it opens a WebSocket connection to the server, keeps track of the current
EFSM state as part of its React component state, and most importantly, renders
the React component encoding of the active EFSM state.
Channel communications are managed by the runtime, which allows it to render
the successor of a receive state upon receiving a message from the channel.
Similarly, the session runtime is responsible for passing the required props
for model transitions to EFSM state React components.
The session runtime component is rendered by the developer and requires, as
props, the \textit{endpoint URL} (so it can open the connection) and a list of
\textit{concrete state components}.

The developer writes their own implementation of each state (mainly to
customise how the state is rendered and inject program logic into state
transitions) by extending the abstract React class components.
The session runtime requires references to these concrete components in order to
render the user implementation accordingly.

\subsubsection{Affine Channel Usage}
A limitation of our browser-side session type encoding is
that channel resources are guaranteed to be used \textit{at most once} as opposed to
\textit{exactly once}.
